# Application: data and model

This chapter describes the empirical analysis that the
[diagnostic chapter](07-application-results.qmd) reads its figure and tables
from: what the survey is, which sample the model is fit to, and how the fit is
produced. The diagnostic itself lives in the next chapter; here we set up the
model behind it. Only the from-scratch application (Track A) touches the
restricted survey files; everything the next chapter shows is read from the two
shipped result objects.

## The survey

The application uses the **2019 National Survey of Early Care and Education
(NSECE)**, conducted by NORC at the University of Chicago under the Office of
Planning, Research, and Evaluation (OPRE), Administration for Children and
Families, U.S. Department of Health and Human Services. The analysis draws on
the restricted-use files, distributed through the Inter-university Consortium
for Political and Social Research as ICPSR Study 38445.

The NSECE is a **stratified, multistage cluster sample**. That design is the
reason the analysis is a natural home for the DER: the sampling weights are
genuinely informative, and the cluster-and-stratum structure is exactly what
the design-based sandwich variance is built from. The restricted files carry
the PSU and stratum identifiers that the design-PSU target needs; the public-use
files do not.

## The analysis sample

The model is fit to center-based providers. The analysis sample is
$N = 6{,}785$ providers nested in $J = 51$ groups --- the 50 states plus the
District of Columbia. Because the groups are states and every state is
represented, all 51 groups are observed **by construction**: there is no
group-sampling stage, which is what makes this a clean instance of the
PSUs-within-groups structure from the [simulation](03-simulation-design.qmd)
rather than the sampled-groups structure. The design has **415 primary sampling
units within 30 strata**. State sample sizes are uneven, ranging from 17 to
1{,}110 providers (median 64).

The Stan-ready metadata --- dimensions, the declared weight convention, the
raw-weight Kish design effect, and the fit seed --- travel with the shipped
result object and can be printed directly:

```{r}
#| label: meta
#| echo: false
refit <- readRDS("../data/precomputed/application/nsece_v2_refit.rds")
meta  <- refit$stan_data_meta
tab <- data.frame(
  Quantity = c("Observations (N)", "Groups / states (J)",
               "Fixed-effect columns (p)", "Weight convention",
               "Raw-weight Kish DEFF", "Fit seed", "MCMC configuration"),
  Value = c(format(meta$N, big.mark = ","),
            as.character(meta$J), as.character(meta$p),
            "global unit-mean (raw weights)",
            formatC(meta$kish_raw, format = "f", digits = 2),
            as.character(meta$seed), meta$mcmc),
  check.names = FALSE)
knitr::kable(tab, row.names = FALSE,
             caption = "NSECE fit metadata, read from `nsece_v2_refit.rds`.")
```

## The model

The outcome is provider participation in the state Infant--Toddler
quality-improvement system, a binary indicator with a weighted prevalence of
about 64.7%. The model is a **survey-weighted hierarchical logistic
regression** with two fixed-effect covariates and a state random intercept,

$$
z_{ij} \mid \theta_j \;\sim\; \text{Bernoulli}\!\left(\text{expit}\big(\beta_0 + \beta_1\, \text{poverty\_cwc}_{ij} + \beta_2\, \text{tiered\_reim}_j + \theta_j\big)\right),
\qquad \theta_j \sim \mathcal{N}(0, \sigma^2_\theta),
$$

fit for $j = 1, \dots, 51$ states as a weighted pseudo-posterior. The two
covariates are chosen to sit on opposite sides of the tier structure:

- **`poverty_cwc`** ([I-a]{.tier .tier-ia}) is the community poverty rate,
  group-mean-centered **within each state**. After centering it carries only
  within-state variation, so $\beta_1$ is a purely within-state covariate ---
  the one coefficient the diagnostic should expose to the design.
- **`tiered_reim`** ([I-b]{.tier .tier-ib}) is a state-level indicator for a
  tiered reimbursement policy. It is constant within each state, so $\beta_2$ is
  a between-state coefficient, shielded by the group structure.
- The **51 state random intercepts** $\theta_j$ ([II]{.tier .tier-ii}) are
  protected by shrinkage.

The parameter vector is therefore the three fixed effects, the 51 random
intercepts, and the group scale $\sigma_\theta$ ([III]{.tier .tier-iii}, a
variance hyperparameter that is not certified) --- **54 parameters** enter the
DER analysis (the intercept, two coefficients, and 51 state effects), matching
the $3 + J = 54$ counted across the next chapter's tables.

The pattern mirrors the simulation exactly: a within-group covariate that should
approximate the design effect, a between-group covariate that should be
shielded, and shrinkage-protected random effects. The difference is that here we
do not know the truth, so the diagnostic is doing real work rather than being
checked against a planted answer.

## Declared choices

A DER is defined only once two choices are declared: the weight-scaling
convention and the aggregation unit (see [the DER in brief](01-der-in-brief.qmd)).
The application fixes them as follows.

**Weight convention.** The headline fit uses the **global unit-mean** convention
on the raw design weights: each raw weight is divided by the overall mean of the
raw weights, so the weights average to one across the full sample. On the raw
weights this convention has a Kish design effect of about 3.76 (a weight
coefficient of variation of about 1.66). The alternative within-state scaling is
carried only for the [sensitivity analysis](07-application-results.qmd) in the
next chapter; because the weights enter the pseudo-likelihood, it is a different
pseudo-posterior, not a reweighting of the same draws.

**Aggregation unit.** Both declared targets are reported. The **design-PSU
target** (415 PSUs within 30 strata) is the primary, design-based answer: it
aggregates the score contributions at the level the survey was actually drawn.
The **model-group target** (51 states) is the lens for the closed forms, since
the tier decomposition is stated in terms of the model group. The two targets
genuinely differ here --- the design PSU is not the model group --- and the next
chapter shows that they give different flag sets.

## How the data are prepared and fit

Three scripts under `code/02_application/` carry the Track A application, run in
order.

`app_01_prepare_data.R` builds the model-ready arrays from the restricted-use
analysis extract. It group-mean-centers the poverty covariate within each state
(and asserts the within-state means are zero to numerical precision), verifies
that the tiered-reimbursement indicator is constant within every state before
treating it as state-level, assembles the design matrix
$X = [\text{intercept}, \text{poverty\_cwc}, \text{tiered\_reim}]$, and forms the
within-state sensitivity weights. It writes two objects into the git-ignored
restricted directory: `phase3_data.rds` (the model arrays `N`, `J`, `p`, `y`,
`X`, `group`, plus `w`, the within-state sensitivity convention) and
`phase3_analysis.rds` (an enriched data frame that carries the raw design weight
together with the PSU and stratum indices the design-PSU target needs). Neither
file is shipped. The headline v2 fit does not use `phase3_data$w`; it rebuilds
its declared global unit-mean weights from `phase3_analysis$weight`.

`app_02_fit_refit.R` fits the model. It forms the declared global unit-mean
weights from the raw column, compiles the Stan model in `stan/hlr_weighted.stan`,
and samples it with `cmdstanr` in the non-centered parameterization: **4 chains
of 2,000 post-warmup draws each (2,000 warmup), with `adapt_delta = 0.95`**. The
fit clears the strict convergence gate --- $\widehat{R} \le 1.004$ on all
parameters, zero divergent transitions, and bulk ESS in the thousands --- and it
supplies a plug-in group scale $\widehat{\sigma}_\theta$ of about 0.65 to the
DER computation. The script then computes the dual-target DER with `svyder`
(`der_compute`, `der_classify`, `der_correct` under both the design-PSU and
model-group targets) and writes the shipped result, `nsece_v2_refit.rds`.

`app_03_der_supplement.R` computes the two quantities that need the microdata
but are not in the headline fit: the weight-convention sensitivity (the DER of
the poverty coefficient and the flag counts under the within-state convention,
evaluated on the stored draws of that fit) and the structural protection factors
$R_k$ of the fixed effects, along with the effective per-state sample sizes under
each convention. It writes the shipped `nsece_supp.rds`. As an integrity check it
also re-derives the declared-convention DERs from the stored draws and confirms
they reproduce the shipped refit to within numerical error.

The strict gate and the plug-in $\widehat{\sigma}_\theta$ can be read straight
from the shipped object:

```{r}
#| label: gate
#| echo: false
g <- refit$gate
fmt <- function(x, d) formatC(x, format = "f", digits = d)
tab <- data.frame(
  Check = c("max split-Rhat (all parameters)", "min bulk ESS",
            "min tail ESS", "divergent transitions",
            "plug-in sigma_theta", "strict gate"),
  Value = c(fmt(g$rhat_max, 4), format(round(g$ess_bulk_min)),
            format(round(g$ess_tail_min)), as.character(g$n_divergent),
            fmt(refit$sigma_theta_hat, 3),
            if (isTRUE(g$pass)) "passed" else "failed"),
  check.names = FALSE)
knitr::kable(tab, row.names = FALSE,
             caption = "Convergence gate and plug-in group scale for the NSECE fit (from `nsece_v2_refit.rds`).")
```

## Data access

The split between the two tracks is sharp. **Only Track A of the application**
needs the restricted NSECE data; **Track B** --- the perspective this guide is
written from --- works entirely from the two shipped result files,
`nsece_v2_refit.rds` and `nsece_supp.rds`, and fits nothing.

For the ICPSR access process (Study 38445) and the exact data contract --- the
names, types, and dimensions of the objects the Track A scripts expect, so a
reader with their own NSECE extract can rebuild the inputs --- see
`data/DATA_ACCESS.md`. In brief: the Track A scripts locate the restricted
derivatives through the `NSECE_RESTRICTED_DIR` environment variable, falling back
to the git-ignored `data/nsece/` directory inside the package. That directory,
along with common microdata extensions, is listed in `.gitignore` so that
respondent-level files cannot enter version control by accident.

**No respondent-level data is shipped with this package.** The two objects under
`data/precomputed/application/` carry posterior draws, DER vectors, and a handful
of scalars and short summaries --- all non-identifiable. The restricted NSECE
microdata are never included, never committed, and governed by the terms of the
Restricted Data Use Agreement.

With the fit in hand, the [next chapter](07-application-results.qmd) computes the
diagnostic against both targets and reads off the figure and tables.
